\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-18 analysis/paper/paper.Rmd
fixed comparison table. and consistent themes or figures

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
eval_colophon <- TRUE

@@

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package allows the user to model the effect of time using splines. Also, by including interaction terms between covariates and time, it also allows users to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

@@

\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\

@@

\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\

@@

\caption{Different features of interest in various survival packages.}

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.5,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray30"),
     case.params = list(size = 0.5),
     base.params = list(size = 0.5)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
  scale_y_continuous(labels = scales::label_number_si())
```

@@

@@

```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}

@@

popTimeData <- popTime(data = bmtcrr, time = "ftime")
plot(popTimeData,
     add.case.series = TRUE,
     # add.base.series = TRUE,
     add.competing.event = TRUE,
     comprisk = TRUE)

@@

ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) +
  # geom_line for smooth curve
  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) +
  # geom_step for step function
  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) +
  facet_grid(Disease ~ .) +
  ylim(c(0, 1)) +
  theme_minimal() +
  theme(legend.position = "top") +
  xlab("Time (in Months)") +
  ylab("Relapse risk")
```

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M"))
```

@@

cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4[1:3])

@@

cowplot::plot_grid(plot_row, legend, rel_widths = c(3, .4))
```

@@

lolliplotDots <- lolliplotDots[-which(is.na(lolliplotDots$estimate)), ]
dwplot(lolliplotDots) +
  theme(
    axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
    strip.text.x = element_blank(),
    strip.background = element_rect(colour = "white", fill = "white"),
    legend.position = c(.855, 0.15), legend.text = element_text(size = 11)
  ) +
  # ggtitle("Coefficient plot") +
  labs(color = "Models") +
  scale_color_manual(values = q4[1:3])
```

@@

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M"))
```

@@

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  scale_color_manual(values = q4)

@@

cowplot::plot_grid(plot_row_reg, legend_reg, rel_widths = c(3, .4))
```

@@

stanford_popTime <- popTime(jasa,
  time = "futime",
  event = "fustat"
)
plot(stanford_popTime)
```

@@

```{r stanford-hazard, eval = FALSE, echo=TRUE}
# Compute hazards---
# First, create a list of time points for both exposure status
hazard_data <- expand.grid(
  exposure = c(0, 1),
  futime = seq(0, 1000,
    length.out = 100
  )
)
# Set the offset to zero
hazard_data$offset <- 0
# Use predict to get the fitted values, and exponentiate to
# transform to the right scale
hazard_data$hazard <- exp(predict(fit4,
  newdata = hazard_data,
  type = "link"
))
# Add labels for plots
hazard_data$Status <- factor(hazard_data$exposure,
  labels = c("NoTrans", "Trans")
)

@@

colnames(absrisk) <- c("Time", "NoTrans", "Trans")

# Rearrange the data
absrisk <- gather(
  as.data.frame(absrisk),
  "Status", "Risk", -Time
)

ggplot(absrisk, aes(Time, Risk, colour = Status)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "top") +
  expand_limits(y = 0.5) +
  xlab("Follow-up time") +
  ylab("Cum. Incidence")
```

@@

>>>>

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package allows the user to model the effect of time using splines. Also, by including interaction terms between covariates and time, it also allows users to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

@@

& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
  scale_y_continuous(labels = scales::label_number_si())
```

@@

```{r, eval = eval_cs1, echo=FALSE}
summary(fits[[2]])
```

@@

```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}

@@

popTimeData <- popTime(data = bmtcrr, time = "ftime")

plot(popTimeData, 
     add.competing.event = TRUE,
     comprisk = TRUE,
     # add.base.series = TRUE,
     # ratio = 0.7,
     # facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     competing.params = list(size = 0.8)) + 
  theme_bw(base_size = 10) + 
  theme(legend.position = "bottom") + 
  scale_y_continuous(labels = scales::label_number_si())

# plot(popTimeData,
#      add.case.series = TRUE,
#      # add.base.series = TRUE,
#      add.competing.event = TRUE,
#      comprisk = TRUE)

@@

ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) +
  # geom_line for smooth curve
  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) +
  # geom_step for step function
  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) +
  facet_grid(Disease ~ .) +
  ylim(c(0, 1)) +
  theme_bw(base_size = 10) + 
  theme(legend.position = "bottom") +
  xlab("Time (in Months)") +
  ylab("Relapse risk")
```

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")
```

@@

cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4[1:3]) + theme_bw(base_size = 10)

@@

cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2)
```

@@

lolliplotDots <- lolliplotDots[-which(is.na(lolliplotDots$estimate)), ]
dwplot(lolliplotDots) +
  theme(
    axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
    strip.text.x = element_blank(),
    strip.background = element_rect(colour = "white", fill = "white"),
    legend.position = c(.855, 0.15), legend.text = element_text(size = 11)
  ) +
  theme_bw(base_size = 10) + 
  theme(legend.position = "bottom") + 
  # ggtitle("Coefficient plot") +
  labs(color = "Models") +
  scale_color_manual(values = q4[1:3])
```

@@

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
  theme_bw(base_size = 10) + 
  theme(legend.position = "bottom")
```

@@

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Survival time (Years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  theme_bw(base_size = 10) + 
  theme(legend.position = "bottom")+
  scale_color_manual(values = q4)

@@

cowplot::plot_grid(plot_row_reg, legend_reg, rel_heights = c(3, .4), nrow = 2)
```

@@

stanford_popTime <- popTime(jasa,
  time = "futime",
  event = "fustat"
)
plot(stanford_popTime,
     ribbon.params = list(fill = "grey50")) + theme_bw(base_size = 10) + theme(legend.position = "bottom") 
```

@@

```{r stanford-hazard, eval = FALSE, echo=TRUE}
# Compute hazards---
# First, create a list of time points for both exposure status
hazard_data <- expand.grid(
  exposure = c(0, 1),
  futime = seq(0, 1000,
    length.out = 100
  )
)

# Set the offset to zero
hazard_data$offset <- 0
# Use predict to get the fitted values, and exponentiate to
# transform to the right scale
hazard_data$hazard <- exp(predict(fit4,
  newdata = hazard_data,
  type = "link"
))
# Add labels for plots
hazard_data$Status <- factor(hazard_data$exposure,
  labels = c("NoTrans", "Trans")
)

@@

class(absrisk)

plot(absrisk,
     id.names = c("NoTrans","Trans")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")

# colnames(absrisk) <- c("Time", "NoTrans", "Trans")
# 
# # Rearrange the data
# absrisk <- gather(
#   as.data.frame(absrisk),
#   "Status", "Risk", -Time
# )
# 
# ggplot(absrisk, aes(Time, Risk, colour = Status)) +
#   geom_line() +
#   theme_minimal() +
#   theme(legend.position = "top") +
#   expand_limits(y = 0.5) +
#   xlab("Follow-up time") +
#   ylab("Cum. Incidence")
```

@@

We can also plot the Hazard ratio:
```{r}
newtime <- quantile(fit4[["originalData"]][[fit4[["timeVar"]]]],
                    probs = seq(0.01, 0.99, 0.01))

# reference category
newdata <- data.frame(exposure = 0,
                      futime = newtime)

plot(fit4,
     type = "hr",
     newdata = newdata,
     var = "exposure",
     increment = 1,
     xvar = "futime",
     ci = F,
     rug = T)
```
